Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-03-10 supplement.tex
Update supplement.tex
addressing loose ends

Low coverage (4-8x) Illumina short read data were generated for the Nama,
Gumuz, Amhara and Oromo populations as part of the African Diversity Reference
Panel (Sanger / Wellcome Trust) \citep{Gurdasani2015-qy,Pagani2015-pz} and
approved through a secondary data analysis agreement for this project. Briefly,
raw reads were aligned to GRCH37 with BWAmem, duplicates
marked with Picard MarkDuplicates, reads were realigned around indels with GATK
RealignerTargetCreater/IndelRealigner followed by BQSR with dbSNP 137.
Contamination checks were performed requiring that FREEMIX $<0.05$;
contamination checks resulted in the elimination of 22 Nama samples. We note
that the high heterozygosity in these genomes both due to inherent genetic
diversity and admixture may have violated base assumptions for this
heterozygosity check. Genomes were then variant called with GATK3.2 Unified
Genotyper \citep{DePristo2011-up} using joint calling across 2,478 individuals within
the ADRP dataset with a minimum base quality of 17. Data were merged with 1000
Genomes Phase 3 \citep{1000_Genomes_Project_Consortium2015-zq} using the union
of sites identified with bcftools isec (-n+1) \citep{Danecek2021-kc} then refined
with Variant Recalibrator with a truth sensitivity threshold of 99.5\%. HapMap
III and dbSNP 138 served as known sites while 1000 Genomes Phase 1 Omni2.5 and
Phase 1 genomic SNP served as the training set. After VR, no batch effects were
observed along PC1 and PC2 for 1000 Genomes vs. ADRP. Phasing on the combined
dataset was performed via SHAPEIT2 \citep{Delaneau2013-aw} and utilized the duoHMM
option for duos and trios. We highlight 82 Nama genomes which are newly
available (unpublished) under accession number EGAD00001006198.
\comment{This makes is sound like these weren't generated for this project.}
Among these 82
Nama samples, we down-sampled the individuals to minimize close relatives and
admixture, such that 44 Nama genomes were retained. 2nd and 3rd degrees
relatives were inferred from reconstructed pedigrees with Omni2.5 SNP array
data. Individuals with $>70\%$ estimated Khoe-San ancestry were retained for
analysis, after partitioning ancestry into $k=6$ clusters with \texttt{ADMIXTURE}
\comment{Justification for k=6}
\citep{Alexander2009-sw} where alternative ancestries represent European, West
African, Near Eastern, and East African gene flow.

@@

\sgcomment{To be expanded. TODO: Brenna}
\begin{itemize}
    \item From the merged dataset of the African Diversity Reference Panel and
        1000 Genomes phase 3 data, subsampled to population included in this study
    \item West Africa: Mende from Sierra Leone (MSL); South Africa: newly sequenced 
        Nama; East Africa: Gumuz (traditionally hunter-gatherer with low levels of 
        Eurasian admixture), Oromo and Amhara (combined as traditionally 
        agriculturalists with large proportion of back-to-Africa Eurasian ancestry); 
        Eurasian: British (GBR)
    \item Combined with the high coverage Vindija neanderthal genome
    \item For running \Relate, we used a larger set of populations. We kept all 
        African 1000 Genomes populations, along with the Nama, Gumuz, Oromo, and
        Amhara, as well as 4 Eurasian populations from 1000 Genomes: GBR, CEU, PJL,
        and CHB
\end{itemize}

@@

We performed an initial round of optimization including all six populations
with a family of models including single- and multiple-stem scenarios as
described in the previous section. We identified parameters that reached
consistent values across all models. These included the timings of recent or
non-central divergences and admixture events:
\begin{itemize}
    \item East/West African population split: 60ka
    \item East African/European split: 50 ka
    \item Neanderthal introgression to Eurasian branch: 45 ka
    \item Neanderthal/human split: 550 ka
    \item Eurasian back-to-African migration: 12 ka
\end{itemize}
When testing multiple variations of the more complex models, we kept these
values fixed (see Table~\ref{tab:fixed_params} for a complete list of all fixed
parameters). This reduced the potential for overfitting the more complex
models, while reducing the computational cost of optimization.

@@

\subsection{What does this mean for expectation of morphology btw branches?}
\label{morpho_diff}
\aprcomment{Tim to fill in, and edit section header}

@@

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-omni-hapmap-comparison.pdf}
    \caption{\sgcomment{$\pi$ in figure are not rendered properly}
        \textbf{Comparison of LD statistics parsed using two recombination maps}
        The recombination map determines genetic distances between pairs of SNPs,
        so using different maps could result in systematic biases in parsed
        statistics if the maps differ significantly. Here, the solid lines are
        statistics parsed using the HapMapII recombination map, and dashed lines
        are using the OMNI-YRI map. Shading represents 95\% confidence intervals
        from the OMNI map. These two maps results in LD statistics that are
        indistinguishable from one another. Notations for each statistic are
        described in section~\ref{sec:supp-ld-stats} and indexed by populations
        (0: Nama, 1: Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
    }
    \label{fig:supp-omni-hapmap-comparison}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-continuous-migration-fits.pdf}
    \caption{
        \textbf{Continuous-migration model fit to LD statistics.}
        Predicted vs observed two-locus statistics as a function of genetic
        distance for the continuous-migration model \sgcomment{with or without migration?}. Each panel represents a different
        set of two-locus statistics. Solid lines represent estimates using the
        continuous-migration model. Dashed lines represent observed statistics.
        Notation and indexing of statistics are described in the
        Fig.~\ref{fig:supp-omni-hapmap-comparison} caption.
    }
    \label{fig:supp-continuous-migration-fits}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-fsts-all-models.pdf}
    \caption{\sgcomment{Need ABCD in figure!}
        \textbf{Predicted \bm{$F_{ST}$} between contemporaneous populations over time.}
        All models (A: single origin, B: continuous migration, C: merger without stem
        migration, D: merger with stem migration) match observed $F_{ST}$ between
        sampled populations. The continuous-migration model predicts that human stem
        populations remain genetically similar, while the merger with and without
        stem migration models predict a period of increased $F_{ST}$ between ancestors
        of modern humans. This is largely due to the very small inferred $N_e$ in one of
        the stem branches, which leads to a rapid increase in differentiation due to
        drift.
    }
    \label{fig:supp-FST-panels}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-h12s-all-models.pdf}
    \caption{\sgcomment{can we sync with text colors? Need ABCD in figure}
        \textbf{Predicted pairwise differentiation between contemporaneous
        populations.} $H_{i,j}$ is the predicted pairwise differentiation
        between two populations per base-pair, using a mutation rate estimated
        from matching predicted to observed pairwise diversity from the
        intergenic data used in the fits (see Sec.~\ref{sec:supp-mutation}).
        The single-origin model (A) shows larger deviations between observed
        and expected differentiation than the models with early human structure
        (B: continuous migration, C: merger without stem migration, D: merger
        with stem migration). In the merger models (C and D), while $F_{ST}$ is
        large between early stems due to the sharp bottleneck in one of those
        branches, the pairwise differences are comparable to the
        continuous-migration model. Stem 1E and Stem 1S have low
        differentiation, as the bottleneck in Stem 1 (which they both split
        from) sharply reduces diversity by the time of the split. This has the
        effect of large $F_{ST}$ but small $H_{i,j}$.
    }
    \label{fig:supp-h12-panels}
\end{figure}

>>>>

Low coverage (4-8x) Illumina short read data were generated for the Nama,
Gumuz, Amhara and Oromo populations as part of the African Diversity Reference
Panel (Sanger / Wellcome Trust) \citep{Gurdasani2015-qy,Pagani2015-pz} and
approved through a secondary data analysis agreement for this project. Briefly,
raw reads were aligned to GRCH37 with BWAmem, duplicates
marked with Picard MarkDuplicates, reads were realigned around indels with GATK
RealignerTargetCreater/IndelRealigner followed by BQSR with dbSNP 137.
Contamination checks were performed requiring that FREEMIX $<0.05$;
contamination checks resulted in the elimination of 22 Nama samples. We note
that the high heterozygosity in these genomes both due to inherent genetic
diversity and admixture may have violated base assumptions for this
heterozygosity check. Genomes were then variant called with GATK3.2 Unified
Genotyper \citep{DePristo2011-up} using joint calling across 2,478 individuals within
the ADRP dataset with a minimum base quality of 17. Data were merged with 1000
Genomes Phase 3 \citep{1000_Genomes_Project_Consortium2015-zq} using the union
of sites identified with bcftools isec (-n+1) \citep{Danecek2021-kc} then refined
with Variant Recalibrator with a truth sensitivity threshold of 99.5\%. HapMap
III and dbSNP 138 served as known sites while 1000 Genomes Phase 1 Omni2.5 and
Phase 1 genomic SNP served as the training set. After VR, no batch effects were
observed along PC1 and PC2 for 1000 Genomes vs. ADRP. Phasing on the combined
dataset was performed via SHAPEIT2 \citep{Delaneau2013-aw} and utilized the duoHMM
option for duos and trios. We highlight 82 Nama genomes which are newly
available (and previously unpublished) under accession number EGAD00001006198.
\comment{This makes is sound like these weren't generated for this project.}
Among these 82 Nama samples, we down-sampled the individuals to minimize
close relatives and admixture, such that 44 Nama genomes were retained.
Second- and third-degrees relatives were inferred from reconstructed
pedigrees with Omni2.5 SNP array data.
Individuals with $>70\%$ estimated Khoe-San ancestry were retained for
analysis, after partitioning ancestry into $k=6$ clusters with \texttt{ADMIXTURE}
\comment{Justification for k=6}
\citep{Alexander2009-sw} where alternative ancestries represent European, West
African, Near Eastern, and East African gene flow.

@@

From the merged dataset of the African Diversity Reference Panel (ADRP,
citet{}) and the \citet{1000_Genomes_Project_Consortium2015-zq} phase 3 data,
we subsampled to the populations included in this study. From the 1000 Genomes
data, these included all populations in the African super-population (ESN, GWD,
LWK, MSL, and YRI) and four non-African populations (CEU, GBR, CHB, and PJL).
From the ADRP, these included the Amhara, Bakiga, Gumuz, Nama, Oromo, Somali,
Nama, and Zulu. This larger set of populations were used in the PCA and
\texttt{Admixture} analyses. The \texttt{Relate} analysis used a slightly
smaller set, discluding the Bakiga, Somali, and Zulu samples.

For the demographic modeling, we further subset to the regional populations,
from West Africa: Mende from Sierra Leone (MSL); South Africa: Nama from South
Africa (newly sequenced); East Africa: Gumuz (traditionally hunter-gatherer
with low levels of Eurasian admixture), Oromo and Amhara (combined as
traditionally agriculturalists with large proportion of back-to-Africa Eurasian
ancestry); and Europe: British (GBR). These were combined with the
high-coverage Vindija Neanderhal genome \citep{Prufer2017-kk}, resulting in 290
total samples used in demographic inference.

@@

We performed an initial round of optimization including all six populations
with a family of models including single- and multiple-stem scenarios as
described in the previous section. We identified parameters that reached
consistent values across all models. These included the timings of recent or
non-central divergences and admixture events:
\begin{itemize}
    \item East/West African population split: 60ka
    \item East African/European split: 50 ka
    \item Neanderthal introgression to Eurasian branch: 45 ka
    \item Neanderthal/human split: 550 ka
    \item Eurasian back-to-African migration: 12 ka
\end{itemize}
When testing multiple variations of the more complex models, we kept these
values fixed (see Table~\ref{tab:fixed-params} for a complete list of all fixed
parameters). This reduced the potential for overfitting the more complex
models, while reducing the computational cost of optimization.

@@

\subsection{Relating model predictions to morphological variation}
\label{sec:morpho_diff}

Here we show how the predictions of our model can be related to morphological
variation in the fossil record.
Let $\Delta^2$ be the expected squared difference in a morphological measurement
between fossil 1 and fossil 2,
\begin{equation}
    \Delta^2=\E\left[\left(\bar{X}_1-\bar{X}_2\right)^2\right]+\E\left[V_1\right]+\E\left[V_2\right],
\end{equation}
where $\bar{X}_1$ and $V_1$ and $\bar{X}_2$ and $V_2$ are measurement means and
variances for the populations from which fossil 1 and fossil 2 are sampled.

If we assume that the trait being measured is evolving neutrally, which seems
reasonable as an approximation based on studies of morphological (cranial)
variation in humans and Neanderthals
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}, then
\begin{equation}
    \E\left[\left(\bar{X}_1-\bar{X}_2\right)^2\right]=4\sigma_m^2\tau_B-2\sigma_m^2\tau_{W_1}-2\sigma_m^2\tau_{W_2},
\end{equation}
\begin{equation}
    \E\left[V_A\right]=\frac{1}{h^2}\sigma_m^2\tau_{W_1},
\end{equation}
and
\begin{equation}
    \E\left[V_B\right]=\frac{1}{h^2}\sigma_m^2\tau_{W_2},
\end{equation}
\aprcomment{switch between 1/2 and A/B - is that intentional, or a typo?}
where $\sigma^2_m$ is the additive genetic variance introduced by mutation
(per zygote per generation), $h^2$ is the (narrow sense) heritability,
$\tau_{W_1}$ and $\tau_{W_2}$ are the average coalescence time of pairs
of alleles from the same population and $\tau_B$ is the average coalescence
time of pairs of alleles from different populations
\citep{Whitlock1999-un,Weaver2016-id}.
Substituting and simplifying gives
\begin{equation}
    \Delta^2=\sigma_m^2 \left( 4\tau_B+\frac{1-2h^2}{h^2} \left( \tau_{W_1}+\tau_{W_2} \right) \right),
\end{equation}
which relates $\Delta^2$ to between and within population coalescence times.

As a further approximation, if we assume that the heritability of the
measurement is close to one half, then
\begin{equation}
    \Delta^2\approx4\sigma_m^2\tau_B,
\end{equation}
which shows that under these assumptions $\Delta^2$ is proportional to $\tau_B$.

Nucleotide diversity, $\pi$, for pairs of individuals sampled from different
populations is also proportional to $\tau_B$.
Therefore, ratios of $\Delta^2$ can be predicted from ratios of $\pi$
for the same comparisons.
For example, a ratio for which $\pi$ for Stem 1 and Stem 2 is the numerator
and $\pi$ for a pair of contemporary human populations is the denominator
can be used to predict $\Delta^2$ for two fossils sampled from each of the
two stems based on $\Delta^2$ for contemporary humans.
In this way, the predictions of our model can be related to morphological
variation in the fossil record.

@@

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-omni-hapmap-comparison.pdf}
    \caption{
        \textbf{Comparison of LD statistics parsed using two recombination maps}
        The recombination map determines genetic distances between pairs of SNPs,
        so using different maps could result in systematic biases in parsed
        statistics if the maps differ significantly. Here, the solid lines are
        statistics parsed using the HapMapII recombination map, and dashed lines
        are using the OMNI-YRI map. Shading represents 95\% confidence intervals
        from the OMNI map. These two maps results in LD statistics that are
        indistinguishable from one another. Notations for each statistic are
        described in section~\ref{sec:supp-ld-stats} and indexed by populations
        (0: Nama, 1: Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
    }
    \label{fig:supp-omni-hapmap-comparison}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-continuous-migration-fits.pdf}
    \caption{
        \textbf{Continuous-migration model fit to LD statistics.}
        Predicted vs observed two-locus statistics as a function of genetic
        distance for the continuous-migration model with migration between stems.
        Each panel represents a different
        set of two-locus statistics. Solid lines represent estimates using the
        continuous-migration model. Dashed lines represent observed statistics.
        Notation and indexing of statistics are described in the
        Fig.~\ref{fig:supp-omni-hapmap-comparison} caption.
    }
    \label{fig:supp-continuous-migration-fits}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-fsts-all-models.pdf}
    \caption{
        \textbf{Predicted \bm{$F_{ST}$} between contemporaneous populations over time.}
        All models (A: single origin, B: continuous migration, C: merger without stem
        migration, D: merger with stem migration) match observed $F_{ST}$ between
        sampled populations. The continuous-migration model predicts that human stem
        populations remain genetically similar, while the merger with and without
        stem migration models predict a period of increased $F_{ST}$ between ancestors
        of modern humans. This is largely due to the very small inferred $N_e$ in one of
        the stem branches, which leads to a rapid increase in differentiation due to
        drift.
    }
    \label{fig:supp-FST-panels}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-h12s-all-models.pdf}
    \caption{\sgcomment{can we sync with text colors?}
        \textbf{Predicted pairwise differentiation between contemporaneous
        populations.} $H_{i,j}$ is the predicted pairwise differentiation
        between two populations per base-pair, using a mutation rate estimated
        from matching predicted to observed pairwise diversity from the
        intergenic data used in the fits (see Sec.~\ref{sec:supp-mutation}).
        The single-origin model (A) shows larger deviations between observed
        and expected differentiation than the models with early human structure
        (B: continuous migration, C: merger without stem migration, D: merger
        with stem migration). In the merger models (C and D), while $F_{ST}$ is
        large between early stems due to the sharp bottleneck in one of those
        branches, the pairwise differences are comparable to the
        continuous-migration model. Stem 1E and Stem 1S have low
        differentiation, as the bottleneck in Stem 1 (which they both split
        from) sharply reduces diversity by the time of the split. This has the
        effect of large $F_{ST}$ but small $H_{i,j}$.
    }
    \label{fig:supp-h12-panels}
\end{figure}
